---
title: "Titanic Survival Prediction: The Triumph of Simplicity"
author: "Antigravity (AI Specialist)"
date: "2025-12-27"
output: 
  html_document:
    theme: readable
    highlight: tango
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Executive Summary

In our quest to maximize predictive accuracy on the Kaggle Titanic dataset, we explored a wide spectrum of methodologies, ranging from simple heuristics to Deep Learning (MLPs) and rigorous Stacked Generalization (Stacking).

**The Result:** Our **V4 Simple Ensemble** (XGBoost + Random Forest + GLMnet) remains the undefeated champion, achieving a leaderboard score of **0.78947**.

**The Lesson:** On small, tabular datasets like Titanic ($N_{train}=891$), **Variance is the enemy.** 
*   **Deep Learning** (V6) proved too data-hungry, overfitting the training noise.
*   **Stacking** (V9) failed because the meta-learner couldn't extract stable signal from the sparse Out-of-Fold predictions.
*   **Heuristic Overrides** (V8 WCG) were too brittle, failing to generalize to edge cases in the test set.

This notebook documents the **V4 Solution**: a masterclass in aggressive feature engineering paired with conservative, diverse statistical modeling.

---

# 2. Data Loading & Preparation

We start by combining Train and Test sets to ensure consistent factor levels for categorical variables.

```{r load_data}
library(caret)
library(dplyr)
library(stringr)
library(xgboost)
library(ranger)
library(glmnet)

# Load raw data
train <- read.csv("train.csv", stringsAsFactors = FALSE)
test  <- read.csv("test.csv",  stringsAsFactors = FALSE)

train_ids <- train$PassengerId
test_ids  <- test$PassengerId

# Combine for preprocessing
test$Survived <- NA
full <- bind_rows(train, test)

cat("Combined Dataset Dimensions:", dim(full))
```

---

# 3. Feature Engineering: The "Secret Sauce"

The Titanic challenge is 80% feature engineering. Our strategy focuses on extracting social structures.

### 3.1 Social Features (Titles & Group Size)
The `Title` proxy is arguably the strongest predictor of survival (Sex + Age + Status). We also map `FamilySize` and `Deck`.

```{r titles}
# Title Extraction
full$Title <- str_extract(full$Name, "[a-zA-Z]+\\.")
full$Title[full$Title %in% c("Mme.")] <- "Mrs."
full$Title[full$Title %in% c("Mlle.", "Ms.")] <- "Miss."
full$Title[full$Title %in% c("Lady.", "Countess.", "Dona.")] <- "Mrs."
full$Title[full$Title %in% c("Capt.", "Col.", "Don.", "Dr.", "Major.", "Rev.", "Sir.", "Jonkheer.")] <- "Rare"
full$Title <- as.factor(full$Title)

# Family Size
full$FamilySize <- full$SibSp + full$Parch + 1

# Deck (Cabin Proxy)
full$Deck <- ifelse(full$Cabin == "", "U", substr(full$Cabin, 1, 1))
full$Deck <- as.factor(full$Deck)
```

### 3.2 Dynamic Imputation
We don't just fill `Age` with the mean. We impute based on `Title` groups (e.g., a "Master" is younger than a "Mr.").

```{r impute}
# Embarked
full$Embarked[full$Embarked == ""] <- "S"

# Fare (Median)
full$Fare[is.na(full$Fare)] <- median(full$Fare, na.rm = TRUE)

# Age (Group Median by Title)
title_age_medians <- tapply(full$Age, full$Title, median, na.rm = TRUE)
full$Age <- ifelse(is.na(full$Age), title_age_medians[full$Title], full$Age)
```

### 3.3 The "Killer Feature": Survival Rates (FSR/TSR)
This is where V4 shines. We calculate the survival rate of a passenger's *group* (Family or Ticket).
*   **Insight:** People in the same group (Ticket) or Family tend to share a fate.
*   **Implementation:** For each passenger, we look at their group members in the **Training Set**. 
    *   If family lived -> Feature = 1.
    *   If family died -> Feature = 0.
    *   Unknown -> Feature = 0.5.

```{r fsr}
# 1. Family Survival Rate
full$Surname <- sapply(full$Name, function(x) strsplit(x, split = ",")[[1]][1])

full$FamilySurvived <- sapply(1:nrow(full), function(i) {
  pid <- full$PassengerId[i]
  surname <- full$Surname[i]
  fare <- full$Fare[i]
  
  # Find family members in TRAIN set
  family <- train[train$Surname == surname & 
                    train$PassengerId != pid &
                    abs(train$Fare - fare) < 5, ]
  
  if (nrow(family) == 0) return(0.5) # No info
  mean(family$Survived, na.rm = TRUE)
})

# 2. Ticket Survival Rate
full$TicketSurvived <- sapply(1:nrow(full), function(i) {
  pid <- full$PassengerId[i]
  ticket <- full$Ticket[i]
  
  group <- train[train$Ticket == ticket & train$PassengerId != pid, ]
  
  if (nrow(group) == 0) return(0.5) # No info
  mean(group$Survived, na.rm = TRUE)
})

# Maximize the signal
full$GroupSurvived <- pmax(full$FamilySurvived, full$TicketSurvived)
```

---

# 4. Modeling: The "Three Pillars"

We ignore complex stacking in favor of a **Diverse Voting Block**.

1.  **XGBoost**: Gradient Boosting. Best at capturing non-linear interactions.
2.  **Random Forest**: Bagging. Best at handling raw features and reducing variance.
3.  **GLMnet**: Regularized Linear Model. Best at establishing a stable baseline (the "sanity check").

### 4.1 Data Finalization
```{r prep_model}
full$Sex <- as.factor(full$Sex)
full$Embarked <- as.factor(full$Embarked)
full$Pclass <- as.factor(full$Pclass)

# Drop identifiers
full_clean <- full %>% select(-PassengerId, -Name, -Ticket, -Cabin, -Surname)

train_final <- full_clean[1:nrow(train), ]
test_final  <- full_clean[(nrow(train) + 1):nrow(full), ]
test_final$Survived <- NULL
train_final$Survived <- as.factor(ifelse(train$Survived == 1, "Yes", "No"))

# Control: 10-Fold CV
myControl <- trainControl(method = "cv", number = 10, classProbs = TRUE, savePredictions = "final")
```

### 4.2 Training (Conservative Tuning)
Notice `max_depth = 3`. We intentionally limit the XGBoost model's complexity. On small data, deep trees memorize; shallow trees learn.

```{r train}
set.seed(42)

# 1. XGBoost (The Specialist)
model_xgb <- train(
  Survived ~ ., data = train_final, method = "xgbTree", trControl = myControl,
  tuneGrid = expand.grid(nrounds=100, max_depth=3, eta=0.1, gamma=0, 
                         colsample_bytree=0.8, min_child_weight=1, subsample=0.8),
  verbose = FALSE
)

# 2. Random Forest (The Generalist)
model_rf <- train(
  Survived ~ ., data = train_final, method = "ranger", trControl = myControl,
  tuneGrid = expand.grid(mtry=3, splitrule="gini", min.node.size=5),
  importance = "impurity"
)

# 3. GLMnet (The Stabilizer)
model_glm <- train(
  Survived ~ ., data = train_final, method = "glmnet", trControl = myControl,
  tuneGrid = expand.grid(alpha=0.5, lambda=0.01),
  preProcess = c("center", "scale")
)

cat("CV Accuracies:\n")
cat("XGB:", max(model_xgb$results$Accuracy), "\n")
cat("RF: ", max(model_rf$results$Accuracy), "\n")
cat("GLM:", max(model_glm$results$Accuracy), "\n")
```

---

# 5. Ensembling: Democracy Wins

Why averaging? 
*   XGBoost might overfit a specific noise pattern.
*   Random Forest might miss a subtle interaction.
*   GLM might be too linear.
*   **Average**: The errors cancel out.

```{r ensemble}
# Predict Probabilities
p_xgb <- predict(model_xgb, newdata = test_final, type = "prob")$Yes
p_rf  <- predict(model_rf, newdata = test_final, type = "prob")$Yes
p_glm <- predict(model_glm, newdata = test_final, type = "prob")$Yes

# Simple Averaging
final_prob <- (p_xgb + p_rf + p_glm) / 3
final_class <- ifelse(final_prob > 0.5, 1, 0)
```

# 6. Conclusion

This V4 approach is our best because it respects the **Law of Parsimony**. It uses powerful features (`GroupSurvived`) but processes them with conservative, stable models. It resists the temptation to "hack" the leaderboard with fragile overrides, resulting in a model that generalizes well to the unseen private test set.

```{r save}
submission <- data.frame(PassengerId = test_ids, Survived = final_class)
# write.csv(submission, "submission_champion_v4.csv", row.names = FALSE)
head(submission)
```
